Negative Selection in the 20th Century

### 2020-05-19 why-negative-selection.Rmd
clean up graph styling. Create n_older_siblings and fath_age_birth/moth_age_birth. Run regressions with controls for these.

* Divide siblings regressions by own YOB, into early and late.
  - Use this to confirm the basic "selection decreasing" story
  - Actually almost no change.
  - Overall, diffs between siblings and children regressions
    now look more like ascertainment bias than change over time.
  - (I checked my code: I hadn't been controlling for one main term when
    running the interaction. Doh.)

* Plot average scores over time, as an intro graph. DONE
  - Can we work out how to plot selection effects on the "same scale"?
  - Simplest thing would be to use birth dates of children. (Maybe average of
    first and last live birth)
  - Also weight?

* How much does age at first live birth "account for" selection?
  - Need f.3872 age at flb for women with one child

* Work more on weighting the data?

@@

theme_set(theme_minimal())
pval_palette <- c("FALSE" = "grey50", "TRUE" = "orange2")
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dashed")
order_abs <- function (x) x[1] # for use when reordering factors

@@

@@

@@

drake::loadd(famhist)
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

@@

```{r plot-means-over-time, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name),) %>% 
      ggplot(aes(YOB, score, group = score_name)) + 
      geom_line() +
      facet_wrap(~score_name) +
      theme(strip.text = element_text(size = 7)) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")
```

@@

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores are plotted on different scales.", fig.height = 7}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(x = term, y = estimate, shape = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_hline + 
      geom_point(size = 2.5, alpha = 0.8) + 
      labs(y = "Effect size", x = "") +
      facet_wrap(vars(dep.var), ncol = 2, scales = "free_x") +
      coord_flip() +
      scale_colour_manual(values = pval_palette) +
      scale_shape_manual(values = c("circle", "triangle"), 
        guide = guide_legend(title = "")) +
      theme(
        legend.position = "bottom", 
        legend.margin = margin(2,2,2,2)
      )

@@

res_pcs %>% 
      mutate(
        estimate = abs(estimate),
        dep.var  = fct_relevel(dep.var, "N children")
      ) %>% 
      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
        geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
        labs(x = "Effect size (log scale)", y = "") + 
        scale_colour_manual(values = pval_palette)

@@

max_abs <- function (x) max(abs(x))
res_period %>% 
      mutate(
        dep.var       = ifelse(children, "Children", "Siblings"),
        period        = (period == "late") + children,
        score_name    = pretty_names(score_name),
        score_name    = fct_reorder(score_name, estimate, max_abs, 
                          .desc = TRUE),
        p             = diff.p.value
      ) %>% 
      arrange(score_name) %>% 
      slice(1:48) %>% 
ggplot(aes(period, estimate, colour = dep.var, group = dep.var)) + 
      geom_hline(yintercept = 0, colour = "grey30") + 
      geom_point(size = 1.5) +
      geom_line(aes(linetype = p < 0.05/{{n_tests}})) +
      facet_wrap(vars(score_name), nrow = 4) + 
      scale_fill_manual(values = pval_palette) +
      scale_colour_manual(values = c("darkgreen", "navy")) +
      scale_linetype_manual(values = c("FALSE" = "dashed", "TRUE" = "solid"))

@@

res_weighted_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate_weighted, order_abs)
      ) %>% 
      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
        names_to = "Regression", values_to = "estimate") %>% 
      mutate(
        Regression = sub("estimate_", "", Regression)
      ) %>% 
      ggplot(aes(x = term, y = estimate, fill = Regression,
        colour = p.value < 0.05/{{n_regs}})) +
      my_hline + 
      geom_point(size = 1.5, stroke = .8, shape = 21) + 
      labs(y = "Effect size", x = "") +
      coord_flip() +
      scale_fill_manual(values = 
        c("unweighted" = "transparent", "weighted" = "darkgreen")) +
      scale_colour_manual(values = pval_palette)

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings")
      ) %>% 
      ggplot(aes(term, estimate, color = p.value < 0.05/{{n_regs}}, shape  = PCs)) + 
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
        my_hline +
        facet_wrap(vars(dep.var), scales = "free_x") +
        coord_flip() +
        scale_color_manual(values = pval_palette) +
        theme(legend.position = "bottom")
```

@@

res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs)
      ) %>% 
      ggplot(aes(term, estimate, group = sex, colour = sex)) +
        my_hline +
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.7) +
        coord_flip() +
        scale_colour_brewer(palette = 2, type = "qual")

@@

res_combined %>% 
      mutate(
        score_name = fct_reorder(score_name, estimate, order_abs),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        geom_point(shape = 21, size = 1.8, stroke = 0.8) + 
        scale_color_manual(values = pval_palette) +
        scale_fill_brewer(type = "qual", palette = 3)
```

@@

n_regs <- nrow(res_partners)
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs)
      ) %>% 
      ggplot(aes(score_name, estimate, colour = `N partners`)) + 
        my_hline +
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = .8) +
        facet_wrap(vars(sex)) +
        coord_flip() +
        scale_colour_manual(values = c("darkred", "steelblue")) +
        theme(legend.position = "bottom")

@@

drake::loadd(res_edu)
res_edu %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs),
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
      ) %>% 
      ggplot(aes(term, estimate, colour = `Age left FTE`)) +
        my_hline +
        geom_pointrange(aes(ymin=conf.low, ymax=conf.high), alpha = 0.8,
          position = position_dodge(width = 0.4)) + 
        scale_colour_viridis_d(begin = 0.1, end = 0.85) +
        coord_flip()
```

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs),
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        )
      ) %>% 
      ggplot(aes(term, estimate, fill = Income, colour = p.value < 0.05/{{n_regs}})) +
        my_hline +
        geom_point(alpha = 0.8, shape = 21, size = 1.5, stroke = 0.9) + 
        scale_fill_brewer(direction = -1) +
        scale_colour_manual(values = pval_palette) +
        coord_flip()
```

>>>>

* Divide siblings regressions by own YOB, into early and late.
  - Use this to confirm the basic "selection decreasing" story. DONE.
  - Actually almost no change.
  - Overall, diffs between siblings and children regressions
    now look more like ascertainment bias than change over time.
  - (I checked my code: I hadn't been controlling for one main term when
    running the interaction. Doh.)

* Work more on weighting the data?
  - weight by age at first live birth
  - education using ISCED

* direct regressions of scores on age_flb

@@

@@

theme_set(theme_minimal())
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey50", "TRUE" = "orange2")

pval_scale <- function (...) scale_color_manual(values = pval_palette, ...)

pval_geom_point <- function (size = 1.5, stroke = 0.7, ...) {
  geom_point(shape = 21, size = size, stroke = stroke, ...)
}

my_fill_scale <- function (n, aesthetics = "fill", ...) {
  if (n == 2) {
    scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

# for use when reordering factors. Order by the nth score in a group
order_abs <- function (n = 1) {
  function (x) x[n] 
}

@@

  acronyms <- c("WHR", "BMI", "DPW", "CPD", "SI", "AI", "WC")
  for (acronym in acronyms) {
    regex <- paste0("\\b", acronym, "\\b")
    names <- gsub(regex, toupper(acronym), names, ignore.case = TRUE)
  }

@@

```{r calc-means-over-time}

drake::loadd(famhist)
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(
        -YOB, 
        names_to = "score_name",
        values_to = "score"
      )

@@

```{r calc-means-over-time-children, eval = FALSE}
pgs_over_time_children <- famhist %>% 
      mutate(
        YOB_kids = (year_flb + year_llb)/2,
        YOB      = santoku::chop(
                     YOB_kids, 
                     brk_left(seq(1960, 1995, 5), close_end = FALSE), 
                     seq(60, 90, 5),
                     extend = FALSE
                   )
      ) %>% 
      filter(! is.na(YOB)) %>% 
      group_by(YOB) %>% 
      summarize(across(all_of(score_names), 
        ~weighted.mean(.x, n_children, na.rm = TRUE)
      )) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

pgs_over_time <- bind_rows(
        Respondents = pgs_over_time,
        Children    = pgs_over_time_children,
        .id         = "Generation"
      )

```{r plot-means-over-time, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, score, group = score_name)) + 
      geom_line() +
      facet_wrap(~score_name) +
      theme(
        strip.text  = element_text(size = 7), 
        axis.text.x = element_text(size = 7)
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")
```

@@

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children.", fig.height = 7}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      labs(x = "Effect size", y = "") +
      facet_wrap(vars(dep.var), ncol = 2) +
      pval_scale() +
      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
      theme(
        legend.position = "bottom", 
        legend.margin   = margin(2, 2, 2, 2)
      )

@@

res_pcs %>% 
      mutate(
        estimate = abs(estimate),
        dep.var  = fct_relevel(dep.var, "N children")
      ) %>% 
      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
        geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
        labs(x = "Effect size (log scale)", y = "") + 
        pval_scale(aesthetics = "colour")

@@

max_abs <- function (x) max(abs(x))
res_period %>% 
      mutate(
        dep.var       = ifelse(children, "Children", "Siblings"),
        period        = (period == "late") + children,
        score_name    = pretty_names(score_name),
        score_name    = fct_reorder(score_name, estimate, max_abs, 
                          .desc = TRUE),
        p             = diff.p.value
      ) %>% 
      arrange(score_name) %>% 
      slice(1:48) %>% 
ggplot(aes(period, estimate, colour = dep.var, group = dep.var)) + 
      geom_hline(yintercept = 0, colour = "grey30") + 
      geom_point(size = 1.5) +
      geom_line(aes(linetype = p < 0.05/{{n_tests}})) +
      facet_wrap(vars(score_name), nrow = 4) + 
      pval_scale() +
      my_fill_scale(n = 2, aesthetics = "colour") +
      scale_linetype_manual(values = c("FALSE" = "dashed", "TRUE" = "solid"))

@@

res_weighted_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate_weighted, order_abs(1))
      ) %>% 
      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
        names_to = "Regression", values_to = "estimate") %>% 
      mutate(
        Regression = sub("estimate_", "", Regression)
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = Regression,
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      labs(x = "Effect size", y = "") +
      my_fill_scale(n = 2) +
      pval_scale(aesthetics = "colour")

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(2)),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings"),
        dep.var = fct_relevel(dep.var, "N Siblings")
      ) %>% 
      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
        fill  = PCs)) + 
        pval_geom_point(alpha = 0.8) +
        my_vline +
        facet_wrap(vars(dep.var)) +
        pval_scale() +
        my_fill_scale(n = 2, direction = -1) +
        theme(legend.position = "bottom")
```

@@

n_regs <- as.double(nrow(res_sex))
res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2, direction = -1)

@@

res_combined %>% 
      mutate(
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3)
```

```{r calc-age-birth-parents}

drake::loadd(res_age_birth_parents)
drake::loadd(res_all)

res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
compare_abp_raw <- res_age_birth_parents %>% 
      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
      left_join(res_sibs_raw, by = "term")

cor_raw_fath <- with(
  compare_abp_raw %>% filter(control == "fath_age_birth"), 
  cor(estimate.x, estimate.y))
cor_raw_moth <- with(
  compare_abp_raw %>% filter(control == "moth_age_birth"), 
  cor(estimate.x, estimate.y))
```

The dataset has no information on fathers' age at birth of their first child.
However, we can calculate this information for the parents' generation,
for the subsets of respondents who reported their mother's or father's age
and who had no elder siblings. We run *sibling* regressions on these subsets,
controlling for either parent's age at their birth.
Figure \@ref(fig:plot-age-birth-parents) shows
the results. Effect sizes are very similar, whether controlling for
father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondent's own generation, effect sizes are positively correlated with
the effect sizes from bivariate regressions (father's age at birth: 
$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).

```{r plot-age-birth-parents}

n_regs <- as.double(nrow(res_age_birth_parents)/2)

res_age_birth_parents %>% 
    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
    mutate(
      term    = pretty_names(term),
      term    = fct_reorder(term, estimate, order_abs(2)),
      Control = ifelse(control == "fath_age_birth", 
                  "Father's age at birth", 
                  "Mother's age at birth"
                )
    ) %>% 
    ggplot(aes(y = term, x = estimate, fill = Control, 
      colour = p.value < 0.05/{{n_regs}})) + 
      pval_geom_point(alpha = 0.8) +  
      my_vline +
      my_fill_scale(2) +
      pval_scale()

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
        colour = p.value < 0.05/{{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex)) +
        my_fill_scale(n = 2) +
        pval_scale() +
        theme(legend.position = "bottom")

@@

drake::loadd(res_edu)
n_regs <- as.double(nrow(res_edu))
res_edu %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
      ) %>% 
      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
        fill = `Age left FTE`)) +
        my_vline +
        pval_geom_point() + 
        pval_scale() +
        my_fill_scale(n = 3, direction = -1)

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        )
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = Income, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point(alpha = 0.8) + 
        my_fill_scale(n = 5, direction = -1) +
        pval_scale()
```
